### 2014-11-07 D1-3.tex
Updated intro and summary.

  The deliverable is accompanied with source code of the components in
  Java, API documentation (javadoc), and pre-compiled packages for
  direct installation and testing on mobile devices.

@@

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities.

Images can contain faces of nearby persons to with whom the citizen is
associated. All this data is highly sensitive to the citizens privacy
and can be used against the citizen if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. The most important legislation on the European Level is the Data
Protection Directive 95/46/EC [1], which is also summarized in the
European Charta of Fundamental Rights [2] of EU citizens:

\begin{quote}
Article 7. Respect for private and family life.

Everyone has the right to respect for his or her private and family
life, home and communications.

Article 8. Protection of personal data.

(1) Everyone has the right to the protection of personal data
concerning him or her.

(2) Such data must be processed fairly for specified purposes and on
the basis of the consent of the person concerned or some other
legitimate basis laid down by law.  Everyone has the right of access
to data which has been collected concerning him or her, and the right
to have it rectified.

(3) Compliance with these rules shall be subject to control by an
independent authority.  The Live+Gov fully consortium complies with
this legislation and is moreover committed to convey and enact the
ethical values inherent in these texts.
\end{quote}

Privacy as Control over Private Data

The ethical aspects of privacy have been the subject of study of many
social scientists and philosophers [3]. Many researchers do focus on
the ways in which privacy can be infringed. Thus they invest a great
amount of work in exploring threats, instead of describing why privacy
is so valuable to us. One scholar which follows an alternative aproach
in the context of digital monitoring is Charles Fried [4]. He
investiages, why we are intuitively so sensitive to violations of our
privacy. For him privacy is not asserted as an intrinsic value by
itself, he rather sated:

@@

Depending on the conversation partner, we change the degree of
information we share if it is a total stranger, a colleague, close
friend or a doctor. With doctors we share information of great
intimacy we do not share with anyone else.

Moreover, we trust those friends to not reveal information about us to
others by respecting their privacy. Trust needs the possibility of
unknown failure. If we would constantly monitor our partners, they
cannot fail unnoticed nor can they willingly share that information
with us.  So privacy, according to Fried, is the foundation of our
core relations. And thus it is valuable, because those relations are
essential to human society.

Fried’s study on the understanding of privacy provided a great
contribution to the research on the same term in philosophy and
computer science [5] and despite the fact his text was published in
1970, he already included technologies to its viewpoint (like location
monitoring) that are particularly relevant to our Context.

Live+Gov Privacy Dashboard gives citizens back the Control

In order to meet the the legal requirements and offer the citiens to
keep a maximum of contol about their personal data, Live+Gov systems
implements a number of privacy protection measure. One of the measures
that is under current development is a Privacy Dashboard (Figure
1). Using this dashboard the citizen will be able to take the
following actions: View and export all data from the citizen that is
currently stored in the Live+Gov system Selectively delete parts of
the stored data Get information about processing applied to this data
Consent or decline selectively to processing steps

View the end poducts of data mining applied to his data in a
comprehensive, graphical way These measures provide the citizen with a
maximal respect for their privacy while at the same time allowing
advanced data processing. Although, abusive exploitation of the data
is still possible, the level of transparency offered by the dashboard
builds trust between the citizen and the service provider.

@@

\chapter{Summary}\label{chap:Summary}

>>>>

This deliverable presents the privacy aware, server based reality
sensing infrastructure for Live+Gov applications.  Privacy protection
is a crucial point for public acceptance of advanced data mining
techniques in eParticipation scenarios.  In order to provide such
protection we perform an extensive analysis of all threats to privacy
and implement a number of measures that address these threats in an
adequate way.

The term "privacy'' itself is highly ambiguous and there is no common
understanding what privacy really is in the literature.  In
Chapter \ref{chap:privacy} we discuss different legal and
philosophical texts and about the subject.  The main relevant
legislation is the EU Data Protection Directive (cf. Section
\ref{EUDIR}). The main ethical treatment is provided by Charles Fried
(cf. Section \ref{sec:ethics}), who defines privacy as the
\emph{control over personal information}. Furthermore, seven concrete
privacy types are introduced (cf. Section \ref{sec:privacytypes}) and
the relation to collected sensor data from mobile phones is made.

On the basis of these foundations perform a thorough analysis of all
risks for the privacy of a citizen. The main difficulty we are
presented with is the selection of sensible measures. There are a lot
of anonymization techniques in the literature and it is easy to jump
ahead and implement a few of them while leaving out massive privacy
gaps at other ends. Therefore a structured approach to this problem
was taken.

We have chosen to use the IT Security Reference Model
\cite{Grimm:ItSecRefModel} (cf. Section \ref{sec:GrimmModel}), that
was developed by our colleague from Prof. Grimm from University of
Koblenz, as a framework to conduct a systematic analysis of all risks
that are posed to the privacy of the citizens. The analysis is carried
out in Section \ref{sec:PPA}. Prof. Grimm also contributed with his
expertise and guidance to the execution of this analysis.

The analysis starts by describing an abstract version of the IT
Systems and the involved actors with their interests. We then describe
vulnerabilities of the IT Systems and conflicts between the interests
of the stakeholders. All threats that are posed to the privacy of the
citizen are caused by a conflict of interest between two actors and
exploit a vulnerability of the system.

We identity six main threats that include ``T1. Insufficent Control
Features'', ``T2. Excessive Data Minig'' and ``T4. Surveillance'' and
asses their risk of occurrence. The biggest threat to the citizens
privacy is posed by Insufficient Control Features of the system. As
soon as collected data of Citizens is stored on the servers, all
control over that data is lost if the service provider does not offer
tools for the citizen to control his data.

These risks are addressed by eight recommendations, that when
implemented provide an adequate protection of the citizens privacy.
One surprising outcome, is that the main privacy threats can be
eliminated by enhancing communication with the citizen and
establishing policies and awareness for privacy aspects inside the
company. Thus, a central point in the protection of the citizens privacy is
a ``Privacy Dashboard'' that gives the citizen a maximum of control
over his data. Using this dashboard he is able to view and export all
data from the citizen that is currently stored in the Live+Gov system,
selectively delete parts of the stored data and get information about
processing applied and view the processing results themselves.

In addition, we have implemented four different anonymization methods
for GPS samples, several security mechanisms for infrastructure and
communication channels, moreover we wrote a concise and comprehensive
privacy policy that informs the citizen about all processing steps.

The deliverable is accompanied with source code of the components in
Java, Python and JavaScript, documentation, and pre-compiled packages
of the executable parts.

@@

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities. Images can contain faces of nearby persons
to with whom the citizen is associated. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. We explain the most relevant legislation in Section
\ref{sec:legal}.

The ethical aspects of privacy have been the subject of study of many
social scientists and philosophers. One scholar which is particularly
relevant in our context is Charles Fried. He investiages, why we are
intuitively so sensitive to violations of our privacy. For him privacy
is not asserted as an intrinsic value by itself, he rather sated:

@@

Fried`s study on the understanding of privacy provided a great
contribution to the research on the same term in philosophy and
computer science and despite the fact his text was published in 1970,
he already included technologies to its viewpoint (like location
monitoring) that are particularly relevant to our Context.  We explain
his theory in \ref{sec:ethics} and follow his definition of privacy in
this document.

In this document we perform a thorough analysis of how to protect the
privacy of a citizen and present several implementations that form the
core of our privacy aware Sensor Data Storage and Mining Infrastructure.

We identify six main threats for the citizens privacy, and derive
eight recommendations that should be followed in order to reduce the
associated risks for hazards.  This analysis form the guideline for
the selection of 9 measures that were implemented in our system and
documented in Chapter \ref{chap:impl}.

@@
